---
title: "LDpred2 @WCPG2023"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 70)
knitr::opts_chunk$set(fig.align = 'center', dev = "svg", out.width = "70%",
                      echo = FALSE, comment = "", fig.width = 5, global.par = TRUE)
ICON_R_PROJECT <- icons::fontawesome$brands$`r-project`
ICON_TRI_EXCL <- icons::fontawesome$solid$`exclamation-triangle`
```

class: title-slide center middle inverse

<br>

# Recent and Future Updates to LDpred2 for Polygenic Scores and Inference

<br>

## WCPG 2023

<br>

## Florian Privé (Aarhus Uni, DK)
### `r icons::icon_style(fill = "white", icons::fontawesome$brands$twitter)` `r icons::icon_style(fill = "white", icons::fontawesome$brands$github)` privefl

---

class: title-slide center middle inverse

# Nothing to disclose

---

```{r, out.width="70%", fig.align='center'}
# knitr::include_graphics("https://flagicons.lipis.dev/flags/4x3/dk.svg")
```

---

## Acknowledgments

<br>

Co-authors:

- Bjarni J. Vilhjálmsson (Aarhus Uni, DK)

- Julyan Arbel (INRIA Grenoble, FR)

- Hugues Aschard (Institut Pasteur, FR)

- Bogdan Pasaniuc (UCLA, CA, USA)

- Clara Albiñana (Aarhus Uni, DK)

<br>

Funding:

- Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath)

- Lundbeck Foundation (Fellowship to Prof. B.J. Vilhjálmsson)

---

class: inverse, center, middle

# Thanks!

<br>

Presentation available at [bit.ly/ldpred2_wcpg2023](https://bit.ly/ldpred2_wcpg2023)

<br>

`r icons::icon_style(fill = "white", icons::fontawesome$brands$twitter)` `r icons::icon_style(fill = "white", icons::fontawesome$brands$github)` privefl

.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan)]
